---
title: "Final data analysis"
author: "SL, LC, KD"
format: html
title-block-banner: true
language:
  en:
    title-block-modified: "Updated"
date: "18 December 2023"
date-modified: today
code-fold: true
editor: visual
engine: knitr---
---

# Introduction

This is an outline of the analyses that will be carried out on the island endemics dataset.

# Method

## Constraints to look out for

Our dependent variable need to be treated carefully:

\(1\) Number of species - this is 'count' data (no fractions). The usual linear regression is not suited to count data. We could use a GLM (if the assumptions are met) or else a Poisson regression with a log-link function.

\(2\) Proportion of endemic species: This takes values between 0 and 1 and is therefore 'compositional' data. The usual linear regression cannot be used for this type of data. We'll be using Beta regression for this.

```{r setting.up,echo=FALSE,message=FALSE,warning=FALSE}
##Setting up----
rm(list=ls(all=TRUE))

### Packages ----
packages <- c("tidyverse","knitr","GGally","DHARMa","betareg","AICcmodavg", "MASS","lmtest")

### Install packages not yet installed----
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

### Packages loading----
invisible(lapply(packages, library, character.only = TRUE))
```

```{r import.data,echo=FALSE,message=FALSE,warning=FALSE}
island_data <- read.csv("C:/Users/sandr/OneDrive/SL Folders research etc/SL_LC/Endemism on islands (Katya)/Data/island_data.csv")
```

### Importing data

```{r import.realdata,echo=FALSE,message=FALSE,warning=FALSE}

island_data <- read.csv("C:/Users/sandr/OneDrive/SL Folders research etc/SL_LC/Endemism on islands (Katya)/data/endemism.csv")



```

### Number of species

Two models will be used: one of them relating the number of species on an island to "area", "topographic complexity" and "proximity" and the other adding another variable, "population density".

#### Poisson and negative binomial regressions for number of species - 4 dependent variables

```{r poisson.real.4dv}

# Poisson regression using four independent variables


mod3 <- glm(spp ~
              area+
              topography+
              proximity+
              population,
            data=island_data,
            family = poisson(link = "log"))

summary(mod3)


# test assumptions
#simulationOutput <- simulateResiduals(fittedModel = mod3)
#plot(simulationOutput)

# negative binomial
mod4dvnb <- glm.nb(spp ~
              area+
              topography+
              proximity+
              population,
            data=island_data)

summary(mod4dvnb)


simulated_res <- simulateResiduals(fittedModel = mod4dvnb)

plot(simulated_res)

testOverdispersion(simulated_res)
testZeroInflation(simulated_res)

```

#### Poisson regression for number of species - 3 dependent variables

```{r poisson.real.3dv}

# Poisson regression using three independent variables

mod4 <- glm(spp ~ 
              area+
              topography+
              proximity,
            data=island_data,
            family = poisson(link = "log"))

summary(mod4)

# test assumptions
#simulationOutput <- simulateResiduals(fittedModel = mod4)
#plot(simulationOutput)

# negative binomial

mod3dvnb <- glm.nb(spp ~ 
              area+
              topography+
              proximity,
            data=island_data,
            )

summary(mod3dvnb)
```

```{r poisson.real.2dv}

# Poisson regression using two independent variables

mod2dv <- glm(spp ~ 
              area+
              topography,
              data=island_data,
            family = poisson(link = "log"))

summary(mod2dv)

# test assumptions
#simulationOutput <- simulateResiduals(fittedModel = mod4)
#plot(simulationOutput)

# negative binomial

mod2dvnb <- glm.nb(spp ~ 
              area+
              topography,
            data=island_data,
            )

summary(mod2dvnb)
```

```{r poisson.real.2dv}

# Poisson regression using two independent variables

mod1dv <- glm(spp ~ 
              topography,
              data=island_data,
            family = poisson(link = "log"))

summary(mod1dv)

# test assumptions
#simulationOutput <- simulateResiduals(fittedModel = mod4)
#plot(simulationOutput)

# negative binomial

mod1dvnb <- glm.nb(spp ~ 
              topography,
            data=island_data,
            )

summary(mod1dvnb)


mod1dvnbb <- glm.nb(spp ~ 
              area,
            data=island_data,
            )

summary(mod1dvnbb)
```

# Poisson regression using three independent variables

mod4 \<- glm(spp \~ area+ topography+ proximity, data=island_data, family = poisson(link = "log"))

summary(mod4)

# test assumptions

#simulationOutput \<- simulateResiduals(fittedModel = mod4) #plot(simulationOutput)

# negative binomial

mod4nb \<- glm.nb(spp \~ area+ topography+ proximity, data=island_data, )

summary(mod4nb)

### Proportion of endemic species

Two models will be used: one of them relating the proportion of endemic species on an island to "area", "topographic complexity" and "proximity" and the other adding another variable, "population density".

#### Proportion of endemic species - 3 dependent variables

```{r beta.regression.real.3dv}

mod5 <- betareg::betareg(prop_endemic ~ 
                           area+
                           topography+
                           proximity,
                         data=island_data)

summary(mod5)

```

#### Proportion of endemic species - 4 dependent variables

```{r beta.regression.real.4dv}

mod6 <- betareg::betareg(prop_endemic ~ 
                           area+
                           topography+
                           proximity+
                           population,
                         data=island_data)

summary(mod6)
```

```{r beta.regression.real.2dv}

mod6a <- betareg::betareg(prop_endemic ~ 
                           area+
                           topography,
                         data=island_data)

summary(mod6a)
AICc(mod6a)

mod6b <- betareg::betareg(prop_endemic ~ 
                                                     topography,
                         data=island_data)

summary(mod6b)
AICc(mod6b)

mod6c <- betareg::betareg(prop_endemic ~ 
                           proximity+
                           topography,
                         data=island_data)


summary(mod6c)
AICc(mod6c)
```

### Finding the best model

```{r AIC.real}
# Find the best fitting model
model.set <- list(mod4dvnb, mod3dvnb,mod2dvnb,mod1dvnb,mod1dvnbb)
model.names <- c("glm_all", "glm_area+topo+prox","glm_area+topo","glm_topo","glm_area")

aictab(model.set, modnames = model.names)
```

### Test nested models

```{r test_nested_models}

# Perform likelihood ratio test
lr_test <- lrtest(mod1dvnb,mod2dvnb,mod3dvnb,mod4dvnb)

# View the results
print(lr_test)
```

# Summary of results

## Number of species

### Including "Population density"

#### GLM for number of species - 4 dependent variables

topography P = 0.0278\*

Null deviance: 3752264 on 7 degrees of freedom

Residual deviance: 153099 on 3 degrees of freedom

#### Poisson regression for number of species - 4 dependent variables

area P \< 0.0001\*\*\*

topography P \< 0.0001\*\*\*

population P = 0.001\*\*\*

Null deviance: 1834.114 on 7 degrees of freedom

Residual deviance: 52.774 on 3 degrees of freedom

### Excluding "Population density"

#### GLM for number of species - 3 dependent variables

area P = 0.05\*

topography P = 0.009\*\*\*

Null deviance: 3752264 on 7 degrees of freedom

Residual deviance: 159209 on 4 degrees of freedom

#### Poisson regression for number of species - 3 dependent variables

area P \< 0.0001\*\*\*

topography P \< 0.0001\*\*\*

Null deviance: 1834.114 on 7 degrees of freedom

Residual deviance: 63.162 on 4 degrees of freedom

## Proportion of endemic species

### Including "Population density"

#### Proportion of endemic species - 4 dependent variables

area P \< 0.0001\*\*\*

topography P = 0.002\*\*

proximity P \< 0.0001\*\*\*

population P = 0.003\*\*

### Excluding "Population density"

#### Proportion of endemic species - 3 dependent variables

topography P \< 0.0001\*\*\*

proximity P = 0.015\*

# Removal of outliers (using only large islands)

The analysis will now be carried out again using only the five large islands and excluding the archipelagos.

```{r import.realdata.filtered,echo=FALSE,message=FALSE,warning=FALSE}

island_data <- read.csv("C:/Users/sandr/OneDrive/SL Folders research etc/SL_LC/Endemism on islands (Katya)/data/endemism.csv")

island_data <- island_data |> filter(island != "malta",island != "tuscan_archipelago",island != "balearics")
#island_data <- island_data |> filter(Island != "tuscan_archipelago")
#island_data <- island_data |> filter(Island != "balearics")

```

### Number of species

Two models will be used: one of them relating the number of species on an island to "area", "topographic complexity" and "proximity" and the other adding another variable, "population density".

#### GLM for number of species - 4 dependent variables

```{r glm.real.4dv.fiveguys}

# GLM using four independent variables

mod1 <- glm(spp ~
              area+
              topography+
              proximity+
              population,
            data=island_data)

summary(mod1)

```

#### GLM for number of species - 3 dependent variables

```{r glm.real.3dv.fiveguys}

# GLM using three independent variables

mod2 <- glm(spp ~ 
              area+
              topography+
              proximity,
            data=island_data)

summary(mod2)

```

#### Poisson regression for number of species - 4 dependent variables

```{r poisson.real.4dv.fiveguys}

# Poisson regression using four independent variables


mod3 <- glm(spp ~
              area+
              topography+
              proximity+
              population,
            data=island_data,
            family = poisson(link = "log"))

summary(mod3)


```

#### Poisson regression for number of species - 3 dependent variables

```{r poisson.real.3dv.fiveguys}

# Poisson regression using three independent variables

mod4 <- glm(spp ~ 
              area+
              topography+
              proximity,
            data=island_data,
            family = poisson(link = "log"))

summary(mod4)

```

### Proportion of endemic species

Two models will be used: one of them relating the proportion of endemic species on an island to "area", "topographic complexity" and "proximity" and the other adding another variable, "population density".

#### Proportion of endemic species - 3 dependent variables

```{r beta.regression.real.3dv.five guys}

mod5 <- betareg::betareg(prop_endemic ~ 
                           area+
                           topography+
                           proximity,
                         data=island_data)

summary(mod5)
```

#### Proportion of endemic species - 4 dependent variables

```{r beta.regression.real.4dv.five guys}

#mod6 <- betareg::betareg(prop_endemic ~ 
                          # area+
                          # topography+
                          # proximity+
                          # population,
                        # data=island_data)

#summary(mod6)

```

## New results

## Number of species

### Including "Population density"

#### GLM for number of species - 4 dependent variables

Not worked out; errors noted.

#### Poisson regression for number of species - 4 dependent variables

area P \< 0.0001\*\*\*

topography P \< 0.0001\*\*\*

population P \< 0.001\*\*\*

Null deviance: 5.7856e+02 on 4 degrees of freedom

Residual deviance: -6.7946e-14 on 0 degrees of freedom

### Excluding "Population density"

#### GLM for number of species - 3 dependent variables

Null deviance: 1393487 on 4 degrees of freedom

Residual deviance: 40488 on 1 degrees of freedom

#### Poisson regression for number of species - 3 dependent variables

area P \< 0.0001\*\*\*

topography P \< 0.0001\*\*\*

Null deviance: 578.558 on 4 degrees of freedom

Residual deviance: 43.292 on 1 degrees of freedom

## Proportion of endemic species

### Including "Population density"

#### Proportion of endemic species - 4 dependent variables

Not worked out - errors recorded

### Excluding "Population density"

#### Proportion of endemic species - 3 dependent variables

area P \< 0.0001\*\*\*

topography P \< 0.0001\*\*\*

proximity P \< 0.0001\*\*\*

# Conclusions

| Model                   | Area   | Topography | Proximity | Population |
|-------------------------|--------|------------|-----------|------------|
| All islands 4dv GLM     |        | \*         | ns        | ns         |
| All islands 4dv Poisson | \*\*\* | \*\*\*     | ns        | \*\*\*     |
| All islands 4dv beta    | \*\*\* | \*\*       | \*\*\*    | \*\*       |
| All islands 3dv GLM     | \*     | \*\*\*     | ns        | not tested |
| All islands 3dv Poisson | \*\*\* | \*\*\*     | ns        | not tested |
| All islands 3dv beta    | ns     | \*\*\*     | \*        | not tested |
| 5 islands 4dv GLM       | NA     | NA         | NA        | NA         |
| 5 islands 4dv Poisson   | \*\*\* | \*\*\*     | ns        | \*\*\*     |
| 5 islands 4dv beta      | NA     | NA         | NA        | NA         |
| 5 islands 3dv GLM       | ns     | ns         | ns        | not tested |
| 5 islands 3dv Poisson   | \*\*\* | \*\*\*     | ns        | not tested |
| 5 islands 3dv beta      | \*\*\* | \*\*\*     | \*\*\*    | not tested |
|                         |        |            |           |            |

: Comparison of models

```{r best.subsets}
endemism3 <- island_data |> dplyr::select(spp,topography,area,proximity,population)

#Load the required packages
library(leaps)
# Load the data
#data(endemism3)
# Compute best subsets regression
models <- regsubsets(spp ~ ., data = endemism3, nvmax = 4)
# Print the summary of the models
summary(models)

plot(models, scale = "adjr2")
```

?
